---
title: "POEM and the Competition: A Head-to-Head"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{POEM and the Competition: A Head-to-Head}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7,
                      fig.height = 4.2)
library(POEM)
set.seed(113)
```

The established test for whether *any* of many candidate mediators is active
is the total-indirect-effect Wald test of Guo et al. (2022), which this
package reports as **HDMM**. POEM's power-enhanced test (**PE**) keeps that
test intact and adds a component that reads the marginal signal from
individual mediators. The two therefore agree by construction whenever the
enhancement is zero, and differ only when it is not.

Under the global null the two coincide, and both hold their size. When
the active mediators all push the outcome the same way, both reject and
the enhancement adds a margin (the article's Figure 1(a)); that
homogeneous case is the easy one, and it is not the one that motivates
the method. The cases that do, mediators acting in different directions,
are where the benchmark loses power and the power-enhanced test does not.
This vignette leads with those, then shows the homogeneous case, and lets
the output speak throughout. (Other comparators named
in the article, namely HILMA, GlobalTest, HDMT, and DACT, live in separate
packages; the closing section shows how to fold them in if you have them
installed.)

# One data set, two verdicts

Consider two data sets with the same exposure, mediators, and sample size.
In the first, the active mediators all push the outcome the same way
(**homogeneous**). In the second, they push in opposite directions and
exactly cancel, so the total indirect effect is zero even though four
mediators are genuinely active (**contrasting**).

```{r blind-spot}
homo <- simulate_mediation_data(n = 200, p = 80, outcome = "continuous",
                                pattern = "homogeneous", c1 = 1)
cont <- simulate_mediation_data(n = 200, p = 80, outcome = "continuous",
                                pattern = "contrasting", c1 = 1)

verdict <- function(d) {
  f <- pe_mediation(d$X, d$Y, d$M, outcome = "continuous")
  c(HDMM = f$value[f$term == "pval_hdmm"], PE = f$value[f$term == "pval_pe"])
}
round(rbind(homogeneous = verdict(homo), contrasting = verdict(cont)), 4)
```

When the mediators agree in sign, both tests reject, the power-enhanced
one more decisively. When the mediators cancel, the benchmark sees a zero
total indirect effect and returns a large p-value, while the
power-enhanced test still detects mediation. Both data sets contain real
mediation; only one test sees it in both.

# Size: neither test trades error for power

A fair comparison starts at the null. With no active mediator (`c1 = 0`) a
test should reject about 5% of the time. The enhancement is designed to
vanish under the null, so PE inherits the benchmark's size rather than
buying power with inflated error:

```{r size}
size <- pe_power_curve(n = 150, p = 50, outcome = "continuous",
                       pattern = "contrasting", c1_grid = 0, n_rep = 60)
size[, c("rejection_hdmm", "rejection_pe")]
```

Both sit near the nominal 0.05 (with 60 replications a rate near 0.05 is
known to within about 0.03). Whatever PE gains later, it does not come
from a looser null.

# Power where it counts: heterogeneous mediation

A signal-strength sweep makes the gap visible. The shared helper below draws
the benchmark and the power-enhanced rejection rates on one plot.

```{r draw-helper}
draw <- function(pc, main) {
  plot(pc$c1, pc$rejection_pe, type = "b", pch = 19, ylim = c(0, 1),
       xlab = expression(c[1]), ylab = "rejection rate", main = main)
  lines(pc$c1, pc$rejection_hdmm, type = "b", pch = 1, lty = 2)
  abline(h = 0.05, col = "grey60", lty = 3)
  legend("topleft", c("PE", "HDMM"), pch = c(19, 1), lty = c(1, 2), bty = "n")
}
grid <- c(0, 0.25, 0.5, 0.75, 1)
```

Two heterogeneous settings, both realistic. On the left, the active
mediators **fully cancel** (the total indirect effect is exactly zero). On
the right, six mediators of **mixed sign** only *partially* cancel, leaving a
small nonzero total indirect effect, the kind of messy signal real data
actually presents. In both the benchmark stays near its size across the
grid while the power-enhanced test rises above it.

```{r dominate, fig.height = 4, fig.alt = "Rejection-rate curves showing PE dominating HDMM under fully-cancelling and partially-cancelling heterogeneous mediation"}
p <- 50
# Mixed-sign mediators that only partially cancel: a small but nonzero total
# indirect effect, where the benchmark is weak and the enhancement is not.
alpha_mix <- c(1, -0.9, 0.8, -0.7, 0.6, -0.5, rep(0, p - 6))
tau_mix   <- c(rep(0.3, 6), rep(0, p - 6))

pc_cancel <- pe_power_curve(n = 150, p = p, outcome = "continuous",
                            pattern = "contrasting", c1_grid = grid, n_rep = 30)
pc_mixed  <- pe_power_curve(n = 150, p = p, outcome = "continuous",
                            c1_grid = grid, n_rep = 30,
                            outcome_args = list(alpha_m = alpha_mix, tau = tau_mix))

op <- par(mfrow = c(1, 2))
draw(pc_cancel, "fully cancelling")
draw(pc_mixed, "partially cancelling")
par(op)
```

At `c1 = 1` the rates are
`r sprintf("%.2f", pc_cancel$rejection_hdmm[5])` against
`r sprintf("%.2f", pc_cancel$rejection_pe[5])` (fully cancelling) and
`r sprintf("%.2f", pc_mixed$rejection_hdmm[5])` against
`r sprintf("%.2f", pc_mixed$rejection_pe[5])` (partially cancelling), from
30 replications, so each carries a Monte Carlo standard error of up to
0.09. The benchmark is built on the total indirect effect; when that
quantity is small or zero despite active mediators, it has little to
detect. The power-enhanced test reads the individual mediators directly.
The gap is larger at the article's design, where the default tuning grid
costs less power (see `?pe_lambda_grid`): at `n = 300`, `p = 500`, and
`c1 = 1` the article's Figure 1(b) shows about 0.15 against 0.76.

# Homogeneous Mediation: Both Tests Work, and the Enhancement Still Helps

The easy case. When every active mediator pushes the same way, the total
indirect effect is large and the benchmark is already powerful; the
enhancement adds a margin rather than a rescue:

```{r homo-curve, fig.height = 4, fig.width = 4.2, fig.alt = "Rejection-rate curves under homogeneous mediation, where HDMM and PE nearly coincide"}
pc_homo <- pe_power_curve(n = 150, p = p, outcome = "continuous",
                          pattern = "homogeneous", c1_grid = grid, n_rep = 30)
draw(pc_homo, "homogeneous")
```

At `c1 = 0.5` the benchmark rejects in
`r sprintf("%.0f", 100 * pc_homo$rejection_hdmm[3])` percent of
replications and the power-enhanced test in
`r sprintf("%.0f", 100 * pc_homo$rejection_pe[3])` percent (the article's
Figure 1(a) prints 0.690 against 0.940 at its design). This is the honest
boundary of the claim: where the benchmark is already strong POEM adds a
margin, and where the benchmark is weak it pulls away.

# The pattern holds for binary and count outcomes

The enhancement is not particular to continuous outcomes. Here is the
rejection rate at a fixed contrasting signal (`c1 = 1`) for all three
outcome models; the benchmark stays near its size (within Monte Carlo
error, about 0.1 at 20 replications) while PE is well above it:

```{r cross-outcome}
at_c1 <- function(outcome, c2) {
  pc <- pe_power_curve(n = 180, p = 50, outcome = outcome,
                       pattern = "contrasting", c1_grid = 1, c2 = c2,
                       n_rep = 20)
  c(HDMM = pc$rejection_hdmm, PE = pc$rejection_pe)
}
round(rbind(continuous = at_c1("continuous", 0.5),
            binary     = at_c1("binary", 1),
            count      = at_c1("count", 0.4)), 3)
```

# Beyond a yes/no: which mediators?

The benchmark answers one global question. The power-enhanced screen also
returns *which* mediators are active, and how reliably. Over repeated
contrasting data sets, it recovers the active set with high precision and
moderate recall, while keeping false positives rare:

```{r identification}
id <- pe_identification_study(n = 150, p = 60, outcome = "continuous",
                              pattern = "contrasting", c1_grid = c(0, 1),
                              n_rep = 30, methods = "Bonferroni")
id
```

The familywise error rate is
`r sprintf("%.2f", id$fwer[id$c1 == 0])` at the null and
`r sprintf("%.2f", id$fwer[id$c1 == 1])` at `c1 = 1`; precision and
recall at `c1 = 1` are `r sprintf("%.2f", id$precision[id$c1 == 1])` and
`r sprintf("%.2f", id$recall[id$c1 == 1])` at this small design (30
replications). At the article's design the supplement prints, for the
contrasting setting at `c1 = 0.6`, a familywise error rate of 0.000 with
precision 0.403 and recall 0.122, which the default grid reproduces. The
benchmark offers no comparable list, because a total-indirect-effect test
of zero points to no mediator at all.

# On real data

The contrast is not only a simulation artifact. On the shipped WHO
health-expenditure data, `summary()` reports, group by group, where the
power-enhanced test finds mediation the benchmark does not:

```{r who}
imr <- WHO_mediation_analysis("imr", groupings = c("global", "region", "income"),
                              lambda_grid = seq(0.1, 5, length.out = 100))
summary(imr)
```

In several regions and income groups the benchmark p-value is large (in
the Western Pacific it is near one), yet the power-enhanced test detects a
specific health-expenditure indicator. These are the cases where mediators
act heterogeneously and the conventional test is blind.

(The small region-by-income cells contain indicators that are constant
within the cell. `WHO_mediation_analysis()` drops those automatically; when
calling `pe_mediation()` on such a subset yourself, pass `drop_constant =
TRUE` to do the same.)

# A balanced scorecard

| Situation | HDMM (benchmark) | PE (POEM) |
|---|---|---|
| Null (no mediation) | correct ~5% size | correct ~5% size |
| Homogeneous mediation | powerful | powerful, with a margin |
| Heterogeneous: fully cancelling | no power | recovers the signal |
| Heterogeneous: partially cancelling | weak | recovers the signal |
| Which mediators are active | not provided | identified, with FWER/FDR control |
| Real-data subgroups | misses several | finds them |

PE holds its size within Monte Carlo error of the nominal level in every
setting the article reports, adds a margin where the benchmark is already
strong, and supplies power where the benchmark has little. That is the
whole of the claim, and the rows above are the evidence for it.

# Folding in other comparators

The article also benchmarks against HILMA (Zhou et al. 2020), GlobalTest
(Djordjilovic et al. 2019), and, for individual-mediator identification,
HDMT (Dai et al. 2022) and DACT (Liu et al. 2022). Those methods live in
their own packages, which POEM does not depend on. If you have them
installed, they slot directly alongside a POEM fit on the same simulated
data. For example:

```{r external, eval = FALSE}
d <- simulate_mediation_data(n = 300, p = 500, outcome = "continuous",
                             pattern = "contrasting", c1 = 0.5)
pe <- pe_mediation(d$X, d$Y, d$M, outcome = "continuous")

# HILMA (package 'freebird'): a debiased-lasso total-mediation-effect test.
hilma_p <- freebird::hilma(d$Y, d$M, d$X)$pvalue

# GlobalTest (Bioconductor package 'globaltest'): a score test of the
# mediator block against the outcome.
gt_p <- globaltest::p.value(globaltest::gt(d$Y, d$M))

data.frame(method = c("HDMM", "PE-HDMM", "HILMA", "GlobalTest"),
           pval = c(pe$value[pe$term == "pval_hdmm"],
                    pe$value[pe$term == "pval_pe"], hilma_p, gt_p))
```

Run on contrasting data, the total-indirect-effect methods (HDMM, HILMA)
behave alike, because they are built on the quantity that cancels, while the
power-enhanced test detects the active mediators, the comparison the
article's figures report in full.

# References

Yu, X., and Kelley, K. (in press). Power Enhancement in High-Dimensional
Heterogeneous Mediation Analysis. *Journal of the American Statistical
Association*.

Guo, X., Li, R., Liu, J., & Zeng, M. (2022). High-dimensional mediation
analysis for selecting DNA methylation loci mediating childhood trauma and
cortisol stress reactivity. *Journal of the American Statistical
Association, 117*(539), 1110--1121.
https://doi.org/10.1080/01621459.2022.2053136

Zhou, R. R., Wang, L., & Zhao, S. D. (2020). Estimation and inference for
the indirect effect in high-dimensional linear mediation models.
*Biometrika, 107*(3), 573--589. https://doi.org/10.1093/biomet/asaa016

Djordjilovic, V., Page, C. M., Gran, J. M., Nost, T. H., Sandanger, T. M.,
Veierod, M. B., & Thoresen, M. (2019). Global test for high-dimensional
mediation: Testing groups of potential mediators. *Statistics in Medicine,
38*(18), 3346--3360. https://doi.org/10.1002/sim.8199

Dai, J. Y., Stanford, J. L., & LeBlanc, M. (2022). A multiple-testing
procedure for high-dimensional mediation hypotheses. *Journal of the
American Statistical Association, 117*(537), 198--213.
https://doi.org/10.1080/01621459.2020.1765785

Liu, Z., Shen, J., Barfield, R., Schwartz, J., Baccarelli, A. A., & Lin, X.
(2022). Large-scale hypothesis testing for causal mediation effects with
applications in genome-wide epigenetic studies. *Journal of the American
Statistical Association, 117*(537), 67--81.
https://doi.org/10.1080/01621459.2021.1914634

# Session Information

```{r session-info}
sessionInfo()
```
