---
title: "Power Enhancement for High-Dimensional Mediation: A Guided Tour"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Power Enhancement for High-Dimensional Mediation: A Guided Tour}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7,
                      fig.height = 4.2)
library(POEM)
```

This vignette is written for a reader who is comfortable with regression
but new to high-dimensional mediation. It explains the problem the
package solves, builds up the idea behind the power-enhanced tests,
then walks through a first analysis, the output, the different outcome
types, a simulation study, and finally the real data application from
the article. You do not need to have read the article first; pointers to
it appear where the details live.

## 1. What is a mediation analysis?

A mediation analysis asks *how* an exposure affects an outcome. Instead
of only asking whether a treatment `X` changes an outcome `Y`, it asks
whether `X` works **through** an intermediate variable `M`, the
*mediator*. The classic picture is a chain:

```
        a                 b
   X  ----->   M   ----->   Y
    \                       /
     \________ c' _________/
            (direct)
```

The effect of `X` on `Y` that travels through `M` is the **indirect
effect**: in a linear model it is the product `a * b` of the
`X -> M` path and the `M -> Y` path. The part of `X`'s effect that does
not go through `M` is the **direct effect** `c'`. Their sum is the
**total effect**.

In modern applications there is rarely a single mediator. Genomics
studies screen thousands of methylation sites; the application later in
this vignette has 57 health-expenditure indicators. When there are many
candidate mediators `M = (M_1, ..., M_p)`, the natural first question is
simply:

> **Is there any active mediator at all?** That is, does `X` affect `Y`
> through *any* of the candidate mediators?

This is a single global hypothesis test, and it is what POEM is built
to perform with high power.

## 2. The problem: total-indirect-effect tests have a blind spot

With many mediators, the conventional approach summarizes them by the
**total indirect effect**, written `beta`: the sum over all mediators of
each one's individual indirect effect. The global test then checks
whether `beta = 0`.

This works well when the mediators all push the outcome the same way.
But consider two active mediators whose individual indirect effects are
equal in size and **opposite in sign**: say `+0.5` through `M_1` and
`-0.5` through `M_2`. Both mediators are genuinely active, yet their
contributions cancel, so the total indirect effect is

```
beta = (+0.5) + (-0.5) = 0.
```

A test built on `beta` sees zero and concludes "no mediation," even
though two mediators are plainly active. This is not a rare edge case:
whenever different pathways move an outcome in opposite directions
(**heterogeneous** mediation, and in the extreme **contrasting**
mediation), the total indirect effect understates the truth and the test
loses power. The article calls this out as the core limitation of
existing high-dimensional mediation tests.

## 3. The idea: a power enhancement component

The power-enhanced (PE) test fixes the blind spot by adding a second
piece that cannot cancel. After a penalized fit selects a small set of
candidate-active mediators, the PE component `J_m` adds up, over those
mediators, the **magnitude** of each one's combined path signal. In code
notation, with `alpha_hat[j]` the estimated `M -> Y` coefficient of
mediator `j`, `Gamma_hat[i, j]` the estimated `X -> M` coefficient from
exposure `i`, and `se_*` their standard errors:

```
J_m = sqrt(p) * sum over exposures i and selected mediators j of
      |alpha_hat[j] / se_alpha[j]| * |Gamma_hat[i, j] / se_Gamma[i, j]|
      * 1{both paths individually significant}
```

Two things matter here. First, the indicator keeps a mediator only when
**both** its `X -> M` path and its `M -> Y` path are individually
significant, so noise mediators do not contribute. Second, because
`J_m` sums **absolute values**, a `+0.5` mediator and a `-0.5` mediator
*reinforce* each other instead of cancelling. The full test statistic is
`M_PE = S_n + J_m`, the benchmark Wald statistic `S_n` (the
total-indirect-effect test) plus the enhancement. Under the global null of no mediation, the screen
rejects every mediator, so `J_m = 0` and `M_PE` behaves exactly like
`S_n` (same chi-square reference distribution, so the same Type I error
rate). Under heterogeneous alternatives, `J_m` grows large and the test
gains power. That is the whole idea: **keep the size, add power exactly
where the old test was blind.**

The same indicator that builds `J_m` also tells you *which* mediators
are active, with familywise error rate control (the default) or false
discovery rate control.

## 4. A first analysis, step by step

The package can simulate data so you can see the method work before
bringing your own. `simulate_mediation_data()` generates the patterns
studied in the article. Here is a **contrasting** setting: a few active
mediators whose effects cancel to a zero total indirect effect.

```{r sim}
set.seed(113)
d <- simulate_mediation_data(n = 200, p = 60, outcome = "continuous",
                             pattern = "contrasting", c1 = 1)
```

The result is a list. The four pieces an analysis needs are the
exposure, outcome, mediators, and (optionally) confounders:

```{r sim-pieces}
dim(d$X)   # exposure: 200 observations, 1 column
length(d$Y)   # outcome: 200 values
dim(d$M)   # mediators: 200 x 60
d$active_mediators   # which mediators are truly active (the ground truth)
d$beta               # the total indirect effect: zero, by construction
```

So four mediators are genuinely active, but the total indirect effect is
zero because their effects cancel. This is precisely the situation that
defeats a total-indirect-effect test. Now run the global test:

```{r first-fit}
fit <- pe_mediation(d$X, d$Y, d$M, outcome = "continuous")
fit
```

Read the two p-values. `pval_hdmm` is the benchmark Wald test on the
total indirect effect; it is large, so that test fails to detect any
mediation (exactly the blind spot from Section 2). `pval_pe` is the
power-enhanced test; it is essentially zero, so it detects the mediation
decisively. The footer lists the individual mediators it flagged as
active. The power enhancement, made concrete.

## 5. Reading the result table

Every POEM test returns a tidy table that prints with sensible rounding
(whole numbers without decimals, p-values to four places) while storing
full precision underneath. The rows are:

| term | meaning |
|---|---|
| `stat_hdmm`, `pval_hdmm` | the benchmark Wald statistic `S_n` and its p-value (the total-indirect-effect test of Guo et al.) |
| `stat_pe`, `pval_pe` | the power-enhanced statistic `M_PE` and its p-value |
| `j_pe` | the power enhancement component `J_m` (the amount added to `S_n`) |
| `total_indirect_effect` | the estimated total indirect effect `beta` |
| `n_active_mediators` | how many individual mediators were identified as active |
| `df` | degrees of freedom of the chi-square reference (the number of exposures) |
| `n_candidate_mediators`, `n_observations` | `p` and `n` |

To pull a number out, index the table like any data frame; the stored
value keeps full precision even though the display rounds:

```{r read}
fit$value[fit$term == "pval_pe"]            # the exact p-value
attr(fit, "active_mediators")               # the active mediator indices
```

## 6. Identifying which mediators are active

The active set is controlled by the `method` argument. The default,
`"Bonferroni"`, controls the familywise error rate: it is conservative,
so when it flags a mediator you can be confident, but it may miss weak
ones. The `"BH"` and `"BY"` options control the false discovery rate
instead and typically recover more mediators.

```{r methods}
des_idx <- function(f) attr(f, "active_mediators")
des_idx(pe_mediation(d$X, d$Y, d$M, outcome = "continuous", method = "Bonferroni"))
des_idx(pe_mediation(d$X, d$Y, d$M, outcome = "continuous", method = "BH"))
```

The target error rate is `error_level` (default 0.05). It is distinct
from the significance level you use to read `pval_pe`; the global test
is insensitive to it.

## 7. Binary and count outcomes

The same construction works for non-continuous outcomes through a link
function. Name the outcome type and the front end dispatches to the
logistic (binary) or Poisson (count) model. The total indirect effect is
then on the link scale (log-odds for binary, log-mean for count).

```{r glm}
set.seed(113)
db <- simulate_mediation_data(n = 250, p = 50, outcome = "binary",
                              pattern = "contrasting", c1 = 1, c2 = 1)
pe_mediation(db$X, db$Y, db$M, outcome = "binary")
```

## 8. A simulation study: size and power

`pe_power_curve()` reproduces the article's Monte Carlo studies. It
sweeps the signal strength `c1`, simulating many data sets at each value
and recording how often each test rejects. At `c1 = 0` (no mediation)
the rejection rate estimates the Type I error rate; at nonzero `c1` it
estimates power. A rate from `n_rep` replications carries a Monte Carlo
standard error of about `sqrt(r * (1 - r) / n_rep)`, so the 40-replication
run below knows a rate near 0.05 to within about 0.03 and a rate near 0.5
to within about 0.08.

```{r power, fig.alt = "Rejection-rate curves for the benchmark and power-enhanced tests"}
set.seed(113)
pc <- pe_power_curve(n = 150, p = 50, outcome = "continuous",
                     pattern = "contrasting", c1_grid = c(0, 0.25, 0.5, 0.75, 1),
                     n_rep = 40)
pc

plot(pc$c1, pc$rejection_pe, type = "b", pch = 19, ylim = c(0, 1),
     xlab = expression(c[1] * " (signal strength)"),
     ylab = "empirical rejection rate",
     main = "Contrasting mediation: PE vs. benchmark")
lines(pc$c1, pc$rejection_hdmm, type = "b", pch = 1, lty = 2)
abline(h = 0.05, col = "grey60", lty = 3)
legend("right", c("power-enhanced", "benchmark Wald"), pch = c(19, 1),
       lty = c(1, 2), bty = "n")
```

At `c1 = 0` both tests reject at about the nominal rate. As the
contrasting signal grows the benchmark rejects in
`r sprintf("%.0f", 100 * pc$rejection_hdmm[5])` percent of replications at
`c1 = 1` and the power-enhanced test in
`r sprintf("%.0f", 100 * pc$rejection_pe[5])` percent. That is the
enhancement at work, but at this small design it is modest, and the
reason is worth knowing. The default tuning grid is the article's grid
rescaled to `n` and `p` (see `?pe_lambda_grid`), and its lower end is set
where the mediators the test names keep their familywise error
guarantee. At `n = 150` and `p = 50` that floor also removes signal. The
review-era grid, 20 values from 0.05 to 1, trades the guarantee for
power on the same simulated data sets:

```{r power-review}
pc_review <- pe_power_curve(n = 150, p = 50, outcome = "continuous",
                            pattern = "contrasting",
                            c1_grid = c(0, 0.25, 0.5, 0.75, 1), n_rep = 40,
                            lambda_grid = seq(0.05, 1, length.out = 20),
                            seed = 113)
pc_review
```

With it the power-enhanced test rejects in
`r sprintf("%.0f", 100 * pc_review$rejection_pe[5])` percent of
replications at `c1 = 1`, but the mediators it names can include null
ones: at the article's design that grid raises the identified set's
familywise error rate from 0.000 to about 0.17. At the article's design
(`n = 300`, `p = 500`, 1000 replications) the default grid gives the
article's Figure 1(b): the power-enhanced test rejects in about 76
percent of replications at `c1 = 1` and the benchmark in about 15
percent, and at `c1 = 0.5` the printed rates are 0.331 and 0.087 (the
reproduction vignette runs those settings). Choose the grid by what the
analysis needs: the default when the identified mediators will be
reported, the wider grid when only the global test matters at a small
design.

## 9. The real application: health spending as a mediator

The package ships the article's empirical data, `WHO_health_mediation`:
a panel of 91 WHO member states over 2000--2021. The exposure is annual
growth in GDP per capita, the candidate mediators are 57
health-expenditure indicators from the WHO Global Health Expenditure
Database, and there are five health outcomes (infant and under-five
mortality, life expectancy, low birthweight, and undernourishment). The
substantive question: **does health spending mediate the relationship
between economic growth and population health?**

```{r who-data}
data(WHO_health_mediation)
dim(WHO_health_mediation)
WHO_health_mediation[1:3, c("country", "region", "income", "year",
                            "gdp_growth", "imr", "gge_gdp")]
```

The 57 mediators are strongly intercorrelated (many are different
normalizations of the same spending), which is exactly the
high-dimensional, correlated-mediator regime the method targets. Their
definitions are in `WHO_indicator_codebook`:

```{r codebook}
head(WHO_indicator_codebook, 4)
```

### Building one analysis design

`WHO_mediation_design()` assembles the exposure, outcome, mediator, and
confounder matrices for a chosen outcome, handling the missing-data
coverage and the confounders (region, income group, and year) for you.
The article standardizes the exposure and the mediators and centers the
outcome, but leaves the confounder indicators unscaled; the call below
does the same by hand and passes `scale = FALSE`.

```{r who-design}
des <- WHO_mediation_design("imr")          # infant mortality rate
c(observations = des$n, mediators = ncol(des$M), confounders = ncol(des$Z))

fit <- pe_mediation(scale(des$X), des$Y - mean(des$Y), scale(des$M),
                    Z = des$Z, outcome = "continuous", scale = FALSE,
                    lambda_grid = seq(0.1, 5, length.out = 100))
active <- des$indicators[attr(fit, "active_mediators")]
WHO_indicator_codebook$description[
  match(active, WHO_indicator_codebook$indicator)]
```

The PE test identifies general government expenditure as a share of GDP
(`gge_gdp`) as the mediator through which economic growth reaches infant
mortality, which is the article's global finding.

### Reproducing the article's table

`WHO_mediation_analysis()` runs the whole set of models the article
reports for one outcome (global, then within each WHO region and income
group), using the article's preprocessing, and returns a table in the
shape of the article's Table for that outcome.

```{r who-analysis}
res <- WHO_mediation_analysis("imr", groupings = c("global", "region"),
                              lambda_grid = seq(0.1, 10, length.out = 40))
res[, c("group", "n_countries", "pval_hdmm", "pval_pe", "active_mediators")]
```

Compare the benchmark and PE columns. In several regions the benchmark
test sees nothing (`pval_hdmm` large) while the PE test rejects
decisively and names the responsible indicator: in the Western Pacific
region, for example, the benchmark p-value is near 1 yet the PE test
detects compulsory health insurance and private expenditure as active
mediators. These are the cases where heterogeneous mediation hides the
signal from the conventional test, and they reproduce the findings in
the article.

`summary()` reads that comparison off the table for you, counting the
groups where the PE test detects mediation the benchmark misses:

```{r who-summary}
summary(res)
```

### Why the preprocessing matters

`WHO_mediation_analysis()` applies the article's preprocessing
internally. Passing the raw design to `pe_mediation()` with the default
`scale = TRUE` standardizes the confounder indicators as well, and that
is a different model: on the same tuning grid the benchmark Wald p-value
moves from about 0.041 to about 0.23, and the penalized selection adds a
second indicator, social health insurance (`shi_che`).

```{r who-raw}
raw <- pe_mediation(des$X, des$Y, des$M, Z = des$Z, outcome = "continuous",
                    lambda_grid = seq(0.1, 5, length.out = 100))
c(article_preprocessing = fit$value[fit$term == "pval_hdmm"],
  raw_matrices = raw$value[raw$term == "pval_hdmm"])
des$indicators[attr(raw, "active_mediators")]
```

Use the article's preprocessing (or `WHO_mediation_analysis()`, which
applies it) whenever the goal is to reproduce its numbers.

## 10. Beyond the global test: the toolkit

The global test is the core, but POEM ships a few conveniences around it.

### A formula interface for data frames

If your variables live in a data frame, `pe_mediate()` takes a formula
for the outcome and exposure, the names of the mediator columns, and
(optionally) confounder names, and calls `pe_mediation()` for you.

```{r formula}
set.seed(113)
d <- simulate_mediation_data(n = 200, p = 60, outcome = "continuous",
                             pattern = "contrasting", c1 = 1)
df <- data.frame(y = d$Y, x = d$X[, 1], d$M)
meds <- grep("^X", names(df), value = TRUE)   # the mediator columns
fit <- pe_mediate(y ~ x, data = df, mediators = meds, outcome = "continuous")
```

### Why each mediator? The per-mediator table

`pe_mediators()` opens up the screening step: for every candidate
mediator the penalized fit selected, it shows the two path statistics,
the two path p-values, the combined screening p-value, and whether the
mediator was flagged active. This is how you see *why* the test reached
its conclusion.

```{r mediators}
pe_mediators(fit)
```

`plot()` shows the same evidence graphically, the screening strength of
each selected mediator with the active ones highlighted.

```{r plot-mediators, fig.alt = "Per-mediator screening evidence for the selected candidate mediators"}
plot(fit)
```

### broom verbs

`tidy()` returns the per-mediator table and `glance()` a one-row summary,
so a POEM result drops into broom and tidyverse workflows.

```{r broom}
glance(fit)
```

### All three multiplicity methods at once

Pass `report_all_methods = TRUE` and `pe_selection()` reports the active
set under Bonferroni (familywise error rate), Benjamini-Hochberg, and
Benjamini-Yekutieli (false discovery rate) side by side, as the
supplement of the article does.

```{r selection}
fit_all <- pe_mediation(d$X, d$Y, d$M, outcome = "continuous",
                        report_all_methods = TRUE)
pe_selection(fit_all)
```

### Planning the sample size

`ss_power_pe_mediation()` plans a study: over a grid of candidate sample
sizes it estimates the power of the PE test by simulation and reports the
smallest sample size that reaches the target power. It runs a Monte Carlo
study, so it is shown here but not evaluated.

```{r ssplan, eval = FALSE}
plan <- ss_power_pe_mediation(outcome = "continuous", pattern = "contrasting",
                              p = 50, n_grid = c(80, 120, 160, 200),
                              target_power = 0.8, n_rep = 100)
plan
attr(plan, "recommended_n")
```

## 11. Bringing your own data

To analyze your own study, arrange:

- `X`: an `n`-by-`q` matrix of exposures (often one column).
- `Y`: a length-`n` outcome (continuous, 0/1, or counts).
- `M`: an `n`-by-`p` matrix of candidate mediators (`p` may exceed `n`).
- `Z` (optional): an `n`-by-`d` matrix of confounders.

then call `pe_mediation(X, Y, M, Z, outcome = ...)`. The defaults
standardize the inputs as the method assumes, choose the tuning
parameter by a high-dimensional information criterion from a grid whose
lower end keeps the identified set's error-rate guarantee, and control
the familywise error rate when identifying active mediators. Everything
returned is an ordinary data frame you can index, save, or plot, and the
print footer reports the tuning parameter chosen.

### When POEM Is the Right Tool, and When It Is Not

The tests assume a linear (or generalized linear) outcome model with a
sparse set of active mediators, standardized inputs, and complete data.
Some practical limits, seen on simulated data during the package's
release audit:

- Sample size. The penalized selection needs `n` large relative to
  `log(p)` and to the signal; with `n` near 100 and `p` in the thousands
  under the contrasting pattern the selection keeps a single mediator and
  the power-enhanced test has no power to add. The default grid's floor
  rises as `n` falls (see `?pe_lambda_grid`), which protects the
  identified set but costs global power at small designs; the section on
  the simulation study above shows the size of that cost.
- Strongly correlated mediators. With an autoregressive correlation of
  0.9 among neighboring mediators the selection keeps one representative
  of a correlated block, so the identified set is unstable from sample to
  sample even when the global test rejects.
- Rare binary outcomes. With about 2 to 3 percent events at `n = 300`
  the penalized logistic fit selects nothing at any grid value; the fit
  is empty and the function says so (see `?pe_mediation`).
- Overdispersed counts. The count model is Poisson; under negative
  binomial overdispersion its size was not inflated at `n = 200`,
  `p = 100`, but about a tenth of the fits were empty.
- Missing values are not handled; complete the data first.

## References

Yu, X., and Kelley, K. (in press). Power Enhancement in
High-Dimensional Heterogeneous Mediation Analysis. *Journal of the
American Statistical Association*.

Fan, J., Liao, Y., & Yao, J. (2015). Power enhancement in
high-dimensional cross-sectional tests. *Econometrica, 83*(4),
1497--1541. https://doi.org/10.3982/ECTA12749

Guo, X., Li, R., Liu, J., & Zeng, M. (2022). High-dimensional mediation
analysis for selecting DNA methylation loci mediating childhood trauma
and cortisol stress reactivity. *Journal of the American Statistical
Association, 117*(539), 1110--1121.
https://doi.org/10.1080/01621459.2022.2053136

## Session Information

```{r session-info}
sessionInfo()
```
