---
title: "Reproducing the Article's Analyses With POEM"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Reproducing the Article's Analyses With POEM}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7,
                      fig.height = 4.2)
library(POEM)
```

This vignette walks through **each analysis in the article** "Power
Enhancement in High-Dimensional Heterogeneous Mediation Analysis"
(in press at the *Journal of the American Statistical
Association*) and reproduces it with POEM. It follows the article's own structure: the
Monte Carlo simulation study (its Section 4, with the identification
study in Section S.4.1 of its supplement) and the empirical data
analysis (its Section 5). The companion vignette *Power Enhancement for
High-Dimensional Mediation: A Guided Tour* introduces the method itself;
here the focus is faithful replication.

**Two things to keep in mind while reading.**

1. *Scope of the comparison.* The article benchmarks the power-enhanced
   tests (PE-HDMM for linear outcomes, PE-HDGMM for generalized outcomes)
   against several competitors: HDMM, HILMA, and GlobalTest. POEM
   implements the proposed PE tests and the **HDMM** benchmark (the
   total-indirect-effect Wald test that the PE component augments). Every
   POEM result therefore reports the central contrast of the article's
   figures and tables, **PE versus HDMM**; HILMA and GlobalTest are
   separate methods outside this package.

2. *Run size.* The article's simulations use `n = 300`, `p = 500`, and
   **1,000 replications** per design point. Reproducing that here would
   take far too long for a vignette, so the live code uses small `n`, `p`,
   and `n_rep`; the curves have the right *shape* but are noisier. Each
   simulation section also shows the exact article settings in an
   `eval = FALSE` block you can run yourself.

```{r seed}
set.seed(113)
```

# Part I. Monte Carlo Simulation (Article Section 4)

The simulation study asks two questions of every test. Under the global
null of no active mediator (`c1 = 0`) the rejection rate estimates the
**Type I error rate**, which should sit near the nominal level, `alpha_level = 0.05`.
Away from the null (`c1 != 0`) the rejection rate estimates
**power**. `pe_power_curve()` sweeps `c1` and returns both the benchmark
(`rejection_hdmm`) and power-enhanced (`rejection_pe`) rejection rates at
each grid point.

A small helper draws an article-style rejection-rate figure for one
design:

```{r plot-helper}
plot_rejection <- function(pc, main) {
  plot(pc$c1, pc$rejection_pe, type = "b", pch = 19, ylim = c(0, 1),
       xlab = expression(c[1] * "  (signal strength)"),
       ylab = "empirical rejection rate", main = main)
  lines(pc$c1, pc$rejection_hdmm, type = "b", pch = 1, lty = 2)
  abline(h = 0.05, col = "grey60", lty = 3)
  legend("right", c("PE (proposed)", "HDMM (benchmark)"),
         pch = c(19, 1), lty = c(1, 2), bty = "n")
}
```

## Section 4.1: Linear Mediation Models

The article studies a univariate exposure with `n = 300` continuous
outcomes and `p = 500` candidate mediators whose noise has an
autoregressive `0.5^|i-j|` covariance. The mediator-exposure coefficient
is `Gamma_x = c1 * tau`, where `tau` is 0.1, 0.2, 0.3, 0.4, 0.5 for the
five mediators with a nonzero outcome coefficient and an independent
N(0, 0.5^2) draw for each of the others, the direct effect is
`c2 = 0.5`, and two outcome-mediator patterns are contrasted:

* **Setting (i), homogeneous:** `alpha_m = (1, 0.8, 0.6, 0.4, 0.2, 0, ...)`,
  so all active mediators push the same way and the total indirect effect
  is `beta = 0.7 * c1` (nonzero). This is the regime where ordinary tests
  already work; the question is whether PE *adds* power.
* **Setting (ii), contrasting:** `alpha_m = (1, -0.5, 0.4, -0.3, 0, ...)`,
  four active mediators whose effects cancel so the total indirect effect
  is `beta = 0` for every `c1`. This is the regime that defeats a
  total-indirect-effect test, and where PE is designed to rescue power.

POEM maps these to `pattern = "homogeneous"` and `pattern = "contrasting"`
in `simulate_mediation_data()` and `pe_power_curve()`. The reduced live
run (the article's Figure for linear models):

```{r lm-curves, fig.height = 4, fig.alt = "Reduced reproduction of the article's linear-model rejection-rate figure for the homogeneous and contrasting settings"}
set.seed(113)
lm_homo <- pe_power_curve(n = 150, p = 50, outcome = "continuous",
                          pattern = "homogeneous",
                          c1_grid = c(0, 0.25, 0.5, 0.75, 1),
                          c2 = 0.5, n_rep = 25)
lm_cont <- pe_power_curve(n = 150, p = 50, outcome = "continuous",
                          pattern = "contrasting",
                          c1_grid = c(0, 0.25, 0.5, 0.75, 1),
                          c2 = 0.5, n_rep = 25)

op <- par(mfrow = c(1, 2))
plot_rejection(lm_homo, "(i) homogeneous")
plot_rejection(lm_cont, "(ii) contrasting")
par(op)
```

At this reduced design (`n = 150`, `p = 50`, 25 replications, so a Monte
Carlo standard error of up to 0.10 per rate) the curves have the
article's shape, not its values. Under the **homogeneous** setting both
tests gain power as the signal grows (at `c1 = 1`, benchmark
`r sprintf("%.2f", lm_homo$rejection_hdmm[5])`, PE
`r sprintf("%.2f", lm_homo$rejection_pe[5])`). Under the **contrasting**
setting the benchmark rejects in
`r sprintf("%.0f", 100 * lm_cont$rejection_hdmm[5])` percent of
replications at `c1 = 1` and the power-enhanced test in
`r sprintf("%.0f", 100 * lm_cont$rejection_pe[5])` percent. The
article's values at its design (`n = 300`, `p = 500`, 1000 replications;
its Figure 1 and the supplement's table of designs) are the converged
numbers a reported analysis deserves:

| Setting | `c1` | HDMM | PE-HDMM |
|---|---|---|---|
| homogeneous | 0 | 0.037 | 0.051 |
| homogeneous | 0.5 | 0.690 | 0.940 |
| contrasting | 0 | 0.036 | 0.044 |
| contrasting | 0.5 | 0.087 | 0.331 |
| contrasting | 1 | about 0.15 (read from Figure 1) | about 0.76 (read from Figure 1) |

The default tuning grid, the article's grid rescaled to the design (see
`?pe_lambda_grid`), reproduces these within Monte Carlo error: 500
replications during the package's release audit gave 0.048 and 0.062,
0.666 and 0.940, 0.036 and 0.048, and 0.096 and 0.326 for the first four
rows. One feature of the article's figure is worth noticing: under the
contrasting setting the benchmark's rate rises above its size at
`c1 = 1`. The penalized fit sometimes keeps only one member of a
cancelling pair, and the total indirect effect through the selected
mediators is then not zero. That is a property of the grid as much as of
the test.

To reproduce the article's figure at its actual resolution (including
the negative half of the `c1` grid, which the symmetric design fills in),
run:

```{r lm-paper-scale, eval = FALSE}
c1_grid <- seq(-1, 1, by = 0.1)
lm_homo_full <- pe_power_curve(n = 300, p = 500, outcome = "continuous",
                               pattern = "homogeneous", c1_grid = c1_grid,
                               c2 = 0.5, n_rep = 1000, cores = 4)
lm_cont_full <- pe_power_curve(n = 300, p = 500, outcome = "continuous",
                               pattern = "contrasting", c1_grid = c1_grid,
                               c2 = 0.5, n_rep = 1000, cores = 4)
```

`pe_simulation_study()` is a convenience wrapper that runs several
patterns at once and stacks the results, which is how the article's
figure panels are produced:

```{r lm-study, eval = FALSE}
study <- pe_simulation_study(n = 300, p = 500, outcome = "continuous",
                             patterns = c("homogeneous", "contrasting"),
                             c1_grid = seq(-1, 1, by = 0.1), n_rep = 1000)
```

### The Real-Data-Motivated Heterogeneous Setting (Supplement)

The supplement adds a third, harder pattern: a setting calibrated to the
DNA-methylation case study of Guo et al. (2022), with **eleven active
mediators of mixed sign** that neither all agree (homogeneous) nor exactly
cancel (contrasting). `simulate_guo_mediation()` generates it from the
calibration constants shipped as [`guo_calibration`]; the total indirect
effect is `guo_calibration$beta_per_c1 * c1`, about `-1.597 * c1`. (The
article reports `-1.5977`, computed from the full-precision Guo
coefficients; the package ships those coefficients rounded to the three
decimals printed in the supplement, which give `-1.597`.)

```{r guo-data}
d <- simulate_guo_mediation(c1 = 1, seed = 113)
dim(d$M)                              # 85 x 1008, the case-study dimensions
d$active_mediators                    # the 11 active loci
round(d$beta, 4)                      # beta_per_c1 * c1, i.e. about -1.597
```

The fit below uses a wide grid, 50 values from 0.1 to 10, as the guided
tour does for this setting. With it both tests reject at `n = 150`, and
the power-enhanced screen recovers several of the mixed-sign loci:

```{r guo-fit}
dn <- simulate_guo_mediation(c1 = 1, n = 150, seed = 113)
fit <- pe_mediation(dn$X, dn$Y, dn$M, outcome = "continuous",
                    lambda_grid = seq(0.1, 10, length.out = 50))
fit[fit$term %in% c("pval_hdmm", "pval_pe", "n_active_mediators"), ]
```

The article's figure sweeps `c1` over `c(0, +/- 0.1, ..., +/- 1)` at the
case-study size `n = 85`, with and without the calibrated confounders,
averaged over 1,000 replications:

```{r guo-paper-scale, eval = FALSE}
rej <- function(c1, confounders) {
  ps <- replicate(1000, {
    d <- simulate_guo_mediation(c1 = c1, confounders = confounders)
    f <- pe_mediation(d$X, d$Y, d$M, Z = d$Z, outcome = "continuous",
                      lambda_grid = seq(0.1, 10, length.out = 50))
    c(f$value[f$term == "pval_hdmm"], f$value[f$term == "pval_pe"])
  })
  rowMeans(ps <= 0.05)
}
sapply(seq(-1, 1, by = 0.1), rej, confounders = FALSE)
```

## Section 4.2: Mediation Models With Generalized Linear Outcomes

The same construction is repeated for non-continuous outcomes through a
link function: a **logistic** model for binary outcomes (PE-HDGMM versus
HDGMM) and a **Poisson** model for counts. Again `n = 300`, `p = 500`,
with a homogeneous and a contrasting pattern apiece; the article fixes
the direct effect at `c2 = 1` for the logistic models and `c2 = 0.4` for
the Poisson models.

Here is the dramatic **contrasting** case for each outcome type (the
panel where PE most clearly rescues power), at reduced size:

```{r glm-curves, fig.height = 4, fig.alt = "Reduced reproduction of the article's logistic and Poisson rejection-rate figures under the contrasting setting"}
set.seed(113)
logit_cont <- pe_power_curve(n = 180, p = 50, outcome = "binary",
                             pattern = "contrasting",
                             c1_grid = c(0, 0.25, 0.5, 0.75, 1),
                             c2 = 1, n_rep = 20)
pois_cont <- pe_power_curve(n = 180, p = 50, outcome = "count",
                            pattern = "contrasting",
                            c1_grid = c(0, 0.25, 0.5, 0.75, 1),
                            c2 = 0.4, n_rep = 20)

op <- par(mfrow = c(1, 2))
plot_rejection(logit_cont, "Logistic, contrasting")
plot_rejection(pois_cont, "Poisson, contrasting")
par(op)
```

As in the article, the benchmark HDGMM has little power against
contrasting effects while PE-HDGMM gains substantial power, all while
holding the size at `c1 = 0` (20 replications here, so each rate carries
a Monte Carlo standard error of up to 0.11). The article's contrasting
rows at `n = 300`, `p = 500`, and `c1 = 0.5` (logistic) or `0.4`
(Poisson) are: logistic, size 0.035 and 0.044, power 0.065 and 0.243;
Poisson, size 0.028 and 0.046, power 0.082 and 0.915 (HDGMM then
PE-HDGMM). For binary and count outcomes the default grid is the
review-era grid, under which the release audit reproduced the Poisson
row (0.060 and 0.905) and the logistic row within Monte Carlo error. The
full article designs, including the homogeneous panels:

```{r glm-paper-scale, eval = FALSE}
c1_grid <- seq(-1, 1, by = 0.1)
# Logistic: c2 = 1
for (pat in c("homogeneous", "contrasting"))
  pe_power_curve(n = 300, p = 500, outcome = "binary", pattern = pat,
                 c1_grid = c1_grid, c2 = 1, n_rep = 1000, cores = 4)
# Poisson: c2 = 0.4
for (pat in c("homogeneous", "contrasting"))
  pe_power_curve(n = 300, p = 500, outcome = "count", pattern = pat,
                 c1_grid = c1_grid, c2 = 0.4, n_rep = 1000, cores = 4)
```

## Supplement Section S.4.1: Identifying Individual Mediators (FWER and FDR)

Beyond the global test, the article's supplement studies how well the
PE machinery **identifies which** individual mediators are active, and
whether it controls the familywise error rate (FWER) or false discovery
rate (FDR). POEM exposes this through the `method` argument
(`"Bonferroni"` for FWER, `"BH"` or `"BY"` for FDR) and the
`report_all_methods` switch, which reports all three side by side as the
supplement does.

For a *single* fit, `pe_selection()` reports the active set under each
method side by side, together with the power-enhanced statistic and
p-value each one produces (the screened set feeds `J_m`, so the methods
can give different global p-values, exactly the PE / PE_BH / PE_BY columns
of the article's extended tables):

```{r identification}
set.seed(113)
d <- simulate_mediation_data(n = 200, p = 60, outcome = "continuous",
                             pattern = "contrasting", c1 = 1)
d$active_mediators                                  # ground truth

fit_all <- pe_mediation(d$X, d$Y, d$M, outcome = "continuous",
                        report_all_methods = TRUE)
pe_selection(fit_all)                               # per-method sets + p-values
```

To reproduce the supplement's identification *study* (its empirical FWER,
FDR, precision, and recall over many replications), `pe_identification_study()`
scores each method's selections against the truth set across a grid of
`c1`, with the supplement's definitions: a mediator counts as truly active
when its outcome coefficient is nonzero, so at `c1 = 0` the five such
mediators are the truth set, `fwer` is the probability of selecting a
mediator outside it, and `recall` is the share of the five that were
selected (pass `truth = "mediation"` for the both-paths definition, under
which nothing is active at `c1 = 0` and `recall` is `NA` there):

```{r id-study}
set.seed(113)
id <- pe_identification_study(n = 150, p = 60, outcome = "continuous",
                              pattern = "contrasting", c1_grid = c(0, 0.5, 1),
                              n_rep = 30)
id
```

At this small design the Bonferroni screen names no false mediator in any
of the 30 replications at any `c1`, and at `c1 = 1` its precision is
`r sprintf("%.2f", id$precision[id$c1 == 1 & id$method == "Bonferroni"])`
with recall
`r sprintf("%.2f", id$recall[id$c1 == 1 & id$method == "Bonferroni"])`
(30 replications; standard error up to 0.09). The article's summary,
"close-to-zero empirical FWER and FDR, high precision, but medium-to-low
recall," describes its own design: at `n = 300`, `p = 500`, and
`c1 = 0.6` the supplement's table prints, for the contrasting setting,
FWER 0.000, FDR 0.000, precision 0.403, and recall 0.122, and for the
homogeneous setting 0.000, 0.000, 0.984, and 0.497. The default grid
reproduces those within Monte Carlo error (500 replications during the
release audit: 0.000, 0.000, 0.406, 0.124 and 0.000, 0.000, 0.972,
0.482). The article-scale study raises `n`, `p`, and `n_rep` to
300 / 500 / 1000:

```{r id-study-paper, eval = FALSE}
pe_identification_study(n = 300, p = 500, outcome = "continuous",
                        pattern = "contrasting", c1_grid = seq(0, 1, by = 0.1),
                        n_rep = 1000, cores = 4)
```

# Part II. Empirical Data Analysis (Article Section 5)

The article asks whether health-care expenditure mediates the
relationship between economic growth and population health across the
world's countries. POEM ships that exact data set, `WHO_health_mediation`:
a country-by-year panel (2000--2021) with

* **Exposure** `X`: annual growth in GDP per capita (World Bank),
* **Mediators** `M`: 57 health-expenditure indicators (WHO Global Health
  Expenditure Database),
* **Outcomes** `Y`: five health outcomes, infant mortality (`imr`),
  under-five mortality (`u5mr`), life expectancy (`leb`), low birthweight
  (`lbw`), and undernourishment (`pou`),

with WHO region and World Bank income group available as the grouping and
confounding variables.

```{r who-data}
data(WHO_health_mediation)
dim(WHO_health_mediation)
table(WHO_health_mediation$region)
```

## Section 5, Table 1: Infant Mortality (IMR)

`WHO_mediation_analysis()` reproduces the article's per-outcome table:
it fits the global model (region, income, and year as confounders), then a
separate model within each WHO region (income and year as confounders) and
each income group (region and year as confounders), and reports the HDMM
and PE p-values for the total indirect effect together with the active
mediators the PE component identifies.

The function uses the two tuning grids of Yu and Kelley (in press): a narrower
`lambda_grid_global = seq(0.1, 5, length.out = 100)` for the global
all-country model and a wider `lambda_grid = seq(0.1, 10, length.out = 100)`
for every subgroup model. Both grids are the defaults, so no explicit
arguments are needed to reproduce the article exactly.

```{r who-imr}
imr <- WHO_mediation_analysis("imr", groupings = c("global", "region", "income"))
imr[, c("group", "n_countries", "pval_hdmm", "pval_pe", "active_mediators")]
```

Read this against the article's IMR table. The headline reproductions:

```{r who-imr-values, include = FALSE}
p_hdmm <- function(g) signif(imr$pval_hdmm[imr$group == g], 2)
```

* **Global (ALL):** general government expenditure as a share of GDP
  (`gge_gdp`) is the active mediator; HDMM is borderline
  (`pval_hdmm` = `r p_hdmm("ALL")`) while PE is essentially zero.
* **SEAR (South-East Asia):** HDMM sees nothing (`pval_hdmm` =
  `r p_hdmm("SEAR")`) yet PE rejects decisively, naming external health
  expenditure per capita (`ext_usd2021_pc`). This is a textbook
  heterogeneous-mediation rescue.
* **WPR (Western Pacific):** HDMM is near 1 (`pval_hdmm` =
  `r p_hdmm("WPR")`) while PE detects compulsory health insurance and
  private expenditure (`chi_che`, `pvtd_gdp`).
* **EUR (Europe):** both tests reject; PE identifies general government
  expenditure as a share of GDP (`gge_gdp`).
* **AFR, AMR:** no active mediation, by either test.

These are precisely the cases the article highlights, where contrasting
or heterogeneous mediation hides the signal from the conventional test and
the PE component recovers it.

The income-group rows are returned in the same call; the strongest signals
appear in the low- and high-income groups, matching the article.

`summary()` digests the whole table, highlighting the groups where the
power-enhanced test detects mediation the benchmark misses (the rows the
article prints in bold):

```{r who-summary}
summary(imr)
```

The supplement also gives an **extended** version of each table that breaks
the power-enhanced result out by multiplicity method. Pass
`full_table = TRUE` to get that layout, with the PE p-value and active
mediators reported separately under Bonferroni (FWER), BH, and BY (FDR):

```{r who-full}
WHO_mediation_analysis("imr", groupings = "global", full_table = TRUE)
```

## The Other Outcomes, and the Region-by-Income Cells

The remaining article outcomes (under-five mortality, life expectancy,
low birthweight, and undernourishment) run identically; only the outcome
code changes:

```{r who-others, eval = FALSE}
for (y in c("u5mr", "leb", "lbw", "pou"))
  print(WHO_mediation_analysis(y, groupings = c("global", "region", "income")))
```

The article also drills into the **region-by-income** cells (for
example, AFR-Low or WPR-High). Build any one cell with
`WHO_mediation_design()`, which assembles the exposure, outcome, mediator,
and confounder blocks for a chosen subset, then call `pe_mediation()`. A
small cell can contain indicators that never vary inside it; passing
`drop_constant = TRUE` drops those zero-variance columns with a warning and
keeps the reported active set numbered against the original mediators
(`WHO_mediation_analysis()` does the same internally):

```{r who-cell, warning = FALSE}
cell <- WHO_mediation_design("imr", region = "AFR", income = "Upper-middle")
c(n_countries= cell$n_countries, mediators = ncol(cell$M))

keep <- apply(cell$M, 2L, sd) > 0
fit <- tryCatch(
  pe_mediation(scale(cell$X), cell$Y - mean(cell$Y),
               scale(cell$M[, keep, drop = FALSE]),
               Z = cell$Z, outcome = "continuous", scale = FALSE,
               lambda_grid = seq(0.1, 5, length.out = 100)),
  error = function(e) NULL)
if (is.null(fit)) "singular covariance (cell too small)" else
  cell$indicators[keep][attr(fit, "active_mediators")]
```

## Why a Mediator? Opening Up the Screening Step

For any fit, `pe_mediators()` shows the evidence the PE component acted on,
the two path statistics, their p-values, the combined screening p-value,
and the active flag for every candidate the penalized fit selected. This
is the per-mediator detail behind the last column of the article's
tables.

```{r who-screen}
des <- WHO_mediation_design("imr")                  # global IMR design
gfit <- pe_mediation(scale(des$X), des$Y - mean(des$Y), scale(des$M),
                     Z = des$Z, outcome = "continuous", scale = FALSE,
                     lambda_grid = seq(0.1, 5, length.out = 100))
pe_mediators(gfit)
```

# Reproducibility Notes

* The global, regional, and income analyses above are deterministic.
  `WHO_mediation_analysis()` uses a tuning grid whose defaults are set to be `lambda_grid_global = seq(0.1, 5, length.out = 100)`
  for the all-country model. Calling the function with no explicit grid
  arguments reproduces the active mediators and p-values shown here, which
  the package's test suite checks as a fixed benchmark.
* The Monte Carlo sections are stochastic. The reduced runs reproduce the
  *shape* of the article's curves; the `eval = FALSE` blocks give the
  exact `n = 300`, `p = 500`, 1,000-replication settings. Set `cores` above
  1 to parallelize and `seed` for exact reproducibility.
* Exact real-data p-values depend on the preprocessing the article
  fixes (standardizing the exposure and mediators, centering the outcome,
  leaving the confounder indicators unscaled, and using treatment contrasts
  with alphabetical reference levels for the region and income factors);
  `WHO_mediation_analysis()` applies all of this for you. See the
  guided-tour vignette for doing it by hand.

# Reference

Yu, X., and Kelley, K. (in press). Power Enhancement in High-Dimensional
Heterogeneous Mediation Analysis. *Journal of the American Statistical
Association*.

# Session Information

```{r session-info}
sessionInfo()
```
